Verifying Hierarchical Cache Coherence Protocols: a theorem-prover based modular, reusable proof

### 2014-02-04 paper/Intuition.tex
some modifications of backgrounda

For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in
the cache tree hierarchy, from Invariants \textit{conservative} and
\textit{localCompatibility}, we can show that for any address $a$,
$c.\state[a] \le p.\state[c][a]$. This is because the ancestor
relation is the reflective-transitive closure of the parent relation,
and Invariants \textit{conservative} and the first part
\textit{localCompatibility} of defines a $\ge$ relation between the
states of a parent and its child.

Any two caches $c_1, c_2$ neither of which are an ancestor of the
other will have a lowest common ancestor $x$ which is an ancestor of
both $c_1$ and $c_2$ and has two distinct children $x_1$ and $x_2$
such that they are the ancestors of $c_1$ and $c_2$ respectively. For
any address $a$, $x.\dstate[x_1][a]$ and $x.\dstate[x_1][a]$ are
compatible by the second part of
Invariant~\ref{localCompatibility}. From this, it straightaway follows
that $x_1.\state[a]$ and $x_2.\state[a]$ are compatible (from
Invariant~\ref{conservative}). From the previous paragraph, it follows
that $c_1$ and $c_2$ are also compatible. This proves
Invariant \textit{nonAncestorCompatibility}.

We prove Invariant \textit{latestValue} by strong induction on
time. Consider two cases: cache $c$ is already \clean{} for an address
$a$ at time $t$ or the contrary.  If \clean{}, then it has the latest
value at time $t$. Since none of its children's directory entries can
be in \Mo{}, all its descendants should be no greater than \Sh{} for
that address (as discussed above). Any non-ancestor of $c$, and
non-descendant of $c$ must be compatible (as discussed above). Since
$c$ is \clean, it is at least in the \Sh{} state, so compatibility
means that non-descendants can not be in \Mo{} state. All of $c$'s
ancestors are non-leaves, so they can not process any store
request. Therefore, no cache save $c$ may have process a store
request, which means that $c$ contains the latest value for address
$a$ at time $t+1$ if it was \clean{} at time $t$.

@@

To prove Invariant \textit{conservative} at some time $t$, consider the case
when a child cache $c$ has sent to or received from the parent $p$ some message
before $t$. Consider the last such message $m$ sent by or received at $c$ at
$t'$. Let's say $m$ is sent by $c$. Then $p$ can not have received any message
from $c$ after sending $m$ in order to not violate Invariant \textit{noCross}.
Instead, if $c$ receives $m$ at $t'$, $p$ will not receive any message after
sending $m$ in order to not violate Invariants \textit{noCross} and
\textit{respFIFO}. Let $m$ be sent or received by $p$ at $t''$. In both cases,
$p.\dstate[c][a]$ at $t''+1$ will be equal to $c.\state[a]$ at $t'+1$ (because
the response messages sync up the next-state according to the transitions in
Figure \ref{trans}). Since $p$ can not receive any message between $t''+1$ and
$t$, it can only upgrade $p.\dstate[c][a]$. Since $c.\state[a]$ does not change
between $t'+1$ and $t$, Invariant \textit{conservative} holds at $t$. If no
message has been sent or received by $c$ before $t$, then $p$
wouldn't have received any message sent by $c$ before $t$ as it means that the
message must be sent by $c$ before $t$. Since the initial states of the
directory and the child's state are both set to \In{}, the same argument holds
(in which the parent can only upgrade), thus maintaining Invariant
\textit{conservative}.

To prove Invariant \textit{localCompatibility}, it is enough to show that a
child $c$ upgrades $c.\state[a]$ on receiving a response for $a$ from its
parent $p$ and $p.\dstate[c][a]$ is downgraded on receiving a response for $a$
from $c$. If this is true, then Invariant \textit{localCompatibility} will not
be violated on receiving a response. The transitions of Figure \ref{trans}
ensures that Invariant \textit{localCompatibility} is not violated when sending
a response.

If the value of the state of a child at the time of sending a response is the
same as the value of the directory for that child at the parent at the time of
receiving the response, it follows straightaway that the parent downgrades its
directory on receiving a response from the child, since a child downgrades on
sending a response and the next-states are synced on receiving the response.
The proof for this follows exactly the same arguments presented above to prove
Invariant \textit{conservative} -- we prove by considering the last message
sent or received at the child before $t$ (and the case when no message has been
sent or received).

Invariant \textit{noTransitWrite} can be proved as follows. Consider a response
sent from a parent $p$ at time $t$ received by its child $c$ at time $t'$. No
other response can be received by the parent at time $t_p$ such that $t+1 \le
t_p \le t'$ to not violate Invariant \textit{noCross}, so $p.\dstate[c][a]$ at
$t+1 \le p.\dstate[c][a]$ at $t'$, since the parent can only upgrade its
directory for $c$. Similarly, no response can be sent by the child at time
$t_c$ such that $t \le t_c \le t'$ for the same reason and hence $c.\state[a]$
at $t \le c.\state[a]$ at $t'$ since the child can only downgrade its state.
This means that during $t_p$ such that $t+1 \le t_p \le t'$, $p.\dstate[c][a]
> \In$ since the parent upgraded its directory for $c$ at $t$. Similarly,
during $t_c$ such that $t \le t_c \le t'$, $c.\state[a] < \Mo$ since the
child upgrades its state at $t'$. Using arguments similar to the proof of
Invariant \textit{latestValue}, we can establish that no cache can be in \Mo{}
state for address $a$ when a response is in transit from $p$ to $c$.
Similarly, we can prove that no cache can be in \Mo{} state for address $a$
when a response is in transit for address $a$ from $c$ to $p$. Since data is in
transit at the same time as when the response is in transit, Invariant
\textit{noTransitWrite} follows.

@@

%$t'+1 = p.\dstate[c][a]$ at $t''+1$ because of Invariant \textit{stateChange}.

@@

%then it would either violate Invariant \textit{noCross}, or the fact that no

@@

%$p.\dstate[a]$. This ensures that Invariant \textit{conservative} is obeyed. We

@@

%Invariant \textit{stateChange} can be proved directly from the atomic

@@

%Invariant \textit{respFIFO} can be enforced by the interconnect network which

@@

%%Invariant \textit{respFIFO} is known as point-to-point FIFO ordering in the

@@

%Proving Invariants \textit{inTransitCondition} and \textit{noTransitEqual},
%needs two further invariants, along with Invariant \textit{stateChange}.

@@

%Invariant \textit{inTransitCondition} can be proved as follows. Whenever a

@@

%response to the child, or it will violate Invariant \textit{noCross}. Using the fact that 

>>>>

For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in
the cache tree hierarchy, from Invariants \fullref{conservative} and
\textit{localCompatibility}, we can show that for any address $a$,
$c.\state[a] \le p.\state[c][a]$. This is because the ancestor
relation is the reflective-transitive closure of the parent relation,
and Invariants \fullref{conservative} and the first part
\textit{localCompatibility} of defines a $\ge$ relation between the
states of a parent and its child.

Any two caches $c_1, c_2$ neither of which are an ancestor of the
other will have a lowest common ancestor $x$ which is an ancestor of
both $c_1$ and $c_2$ and has two distinct children $x_1$ and $x_2$
such that they are the ancestors of $c_1$ and $c_2$ respectively. For
any address $a$, $x.\dstate[x_1][a]$ and $x.\dstate[x_1][a]$ are
compatible by the second part of
Invariant~\ref{localCompatibility}. From this, it straightaway follows
that $x_1.\state[a]$ and $x_2.\state[a]$ are compatible (from
Invariant~\ref{conservative}). From the previous paragraph, it follows
that $c_1$ and $c_2$ are also compatible. This proves
Invariant \fullref{nonAncestorCompatibility}.

We prove Invariant \fullref{latestValue} by strong induction on
time. Consider two cases: cache $c$ is already \clean{} for an address
$a$ at time $t$ or the contrary.  If \clean{}, then it has the latest
value at time $t$. Since none of its children's directory entries can
be in \Mo{}, all its descendants should be no greater than \Sh{} for
that address (as discussed above). Any non-ancestor of $c$, and
non-descendant of $c$ must be compatible (as discussed above). Since
$c$ is \clean, it is at least in the \Sh{} state, so compatibility
means that non-descendants can not be in \Mo{} state. All of $c$'s
ancestors are non-leaves, so they can not process any store
request. Therefore, no cache save $c$ may have process a store
request, which means that $c$ contains the latest value for address
$a$ at time $t+1$ if it was \clean{} at time $t$.

@@

To prove Invariant \fullref{conservative} at some time $t$, consider the case
when a child cache $c$ has sent to or received from the parent $p$ some message
before $t$. Consider the last such message $m$ sent by or received at $c$ at
$t'$. Let's say $m$ is sent by $c$. Then $p$ can not have received any message
from $c$ after sending $m$ in order to not violate Invariant \fullref{noCross}.
Instead, if $c$ receives $m$ at $t'$, $p$ will not receive any message after
sending $m$ in order to not violate Invariants \fullref{noCross} and
\textit{respFIFO}. Let $m$ be sent or received by $p$ at $t''$. In both cases,
$p.\dstate[c][a]$ at $t''+1$ will be equal to $c.\state[a]$ at $t'+1$ (because
the response messages sync up the next-state according to the transitions in
Figure \ref{trans}). Since $p$ can not receive any message between $t''+1$ and
$t$, it can only upgrade $p.\dstate[c][a]$. Since $c.\state[a]$ does not change
between $t'+1$ and $t$, Invariant \fullref{conservative} holds at $t$. If no
message has been sent or received by $c$ before $t$, then $p$
wouldn't have received any message sent by $c$ before $t$ as it means that the
message must be sent by $c$ before $t$. Since the initial states of the
directory and the child's state are both set to \In{}, the same argument holds
(in which the parent can only upgrade), thus maintaining Invariant
\textit{conservative}.

To prove Invariant \fullref{localCompatibility}, it is enough to show that a
child $c$ upgrades $c.\state[a]$ on receiving a response for $a$ from its
parent $p$ and $p.\dstate[c][a]$ is downgraded on receiving a response for $a$
from $c$. If this is true, then Invariant \fullref{localCompatibility} will not
be violated on receiving a response. The transitions of Figure \ref{trans}
ensures that Invariant \fullref{localCompatibility} is not violated when sending
a response.

If the value of the state of a child at the time of sending a response is the
same as the value of the directory for that child at the parent at the time of
receiving the response, it follows straightaway that the parent downgrades its
directory on receiving a response from the child, since a child downgrades on
sending a response and the next-states are synced on receiving the response.
The proof for this follows exactly the same arguments presented above to prove
Invariant \fullref{conservative} -- we prove by considering the last message
sent or received at the child before $t$ (and the case when no message has been
sent or received).

Invariant \fullref{noTransitWrite} can be proved as follows. Consider a response
sent from a parent $p$ at time $t$ received by its child $c$ at time $t'$. No
other response can be received by the parent at time $t_p$ such that $t+1 \le
t_p \le t'$ to not violate Invariant \fullref{noCross}, so $p.\dstate[c][a]$ at
$t+1 \le p.\dstate[c][a]$ at $t'$, since the parent can only upgrade its
directory for $c$. Similarly, no response can be sent by the child at time
$t_c$ such that $t \le t_c \le t'$ for the same reason and hence $c.\state[a]$
at $t \le c.\state[a]$ at $t'$ since the child can only downgrade its state.
This means that during $t_p$ such that $t+1 \le t_p \le t'$, $p.\dstate[c][a]
> \In$ since the parent upgraded its directory for $c$ at $t$. Similarly,
during $t_c$ such that $t \le t_c \le t'$, $c.\state[a] < \Mo$ since the
child upgrades its state at $t'$. Using arguments similar to the proof of
Invariant \fullref{latestValue}, we can establish that no cache can be in \Mo{}
state for address $a$ when a response is in transit from $p$ to $c$.
Similarly, we can prove that no cache can be in \Mo{} state for address $a$
when a response is in transit for address $a$ from $c$ to $p$. Since data is in
transit at the same time as when the response is in transit, Invariant
\textit{noTransitWrite} follows.

@@

%$t'+1 = p.\dstate[c][a]$ at $t''+1$ because of Invariant \fullref{stateChange}.

@@

%then it would either violate Invariant \fullref{noCross}, or the fact that no

@@

%$p.\dstate[a]$. This ensures that Invariant \fullref{conservative} is obeyed. We

@@

%Invariant \fullref{stateChange} can be proved directly from the atomic

@@

%Invariant \fullref{respFIFO} can be enforced by the interconnect network which

@@

%%Invariant \fullref{respFIFO} is known as point-to-point FIFO ordering in the

@@

%Proving Invariants \fullref{inTransitCondition} and \textit{noTransitEqual},
%needs two further invariants, along with Invariant \fullref{stateChange}.

@@

%Invariant \fullref{inTransitCondition} can be proved as follows. Whenever a

@@

%response to the child, or it will violate Invariant \fullref{noCross}. Using the fact that 
